---
title: Analysing simulations
author: Gibran Hemani
---

```{r, echo=TRUE, warning=FALSE, message=FALSE}
library(knitr)
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning=FALSE, dev="png", cache=TRUE)
```


Simulations run using [this model](https://www.draw.io/?state=%7B%22ids%22:%5B%221VqjTd4iH2de7sXI4PhUrxazBnYLduH_w%22%5D,%22action%22:%22open%22,%22userId%22:%22104793251732516248976%22%7D)


```{r}
library(tidyverse)
library(pROC)
load("../results/sim_summary.rdata")
```

There are three classes of simulation

- u1: Genetic variants influence a confounder (u1) of X and Y
- u2: Instruments for X have pleiotropic effects on Y that are mediated by some other trait called u2
- u3: Genetic instruments on X also directly influence a mediator (u3) between X and Y



## Setup data

For each simulation there are some invalid instruments. What is the range of invalidity:

```{r}
simres$prop_invalid_bin <- cut(simres$prop_invalid, breaks=c(0, 0.2, 0.4, 0.6, 0.8, 1), ordered=TRUE)
hist(simres$prop_invalid)
simres$prop_invalid_bin2 <- do.call(rbind, strsplit(as.character(simres$prop_invalid_bin), split=","))[,1] %>% gsub("\\(", "", .) %>% as.numeric
table(simres$prop_invalid_bin)
table(simres$set)
```


## U1

The influence of invalid instruments on the performance of each method in terms of **bias**

```{r}
simresu1 <- subset(simres, set == "U1")
sumu1 <- simresu1 %>% 
	group_by(method, prop_invalid_bin2, set) %>%
	summarise(
		n=n(),
		psig=sum(pval < 1e-3, na.rm=TRUE) / n(),
		bias=sum(pdiff > 0.05, na.rm=TRUE) / n(),
		isq=mean(Isq)
	)

p1 <- sumu1 %>% 
filter(!grepl("Hybrid", method)) %>%
ggplot(aes(y=bias, x=prop_invalid_bin2, group=as.factor(method)), colour=as.factor(method)) +
geom_point(aes(colour=as.factor(method))) +
geom_line(aes(colour=as.factor(method))) +
scale_colour_brewer(type="qual") +
labs(x="Proportion of SNPs with pleiotropic effects", y="Proportion of estimates that are unbiased", colour="Method", shape="Outlier detection", linetype="Outlier detection") +
ylim(c(0,1))
p1
```

Influence on power / false discovery rates. Calculate the AUROC for each method across the simulations


```{r}
sumu1b <- simresu1 %>% 
	group_by(method, prop_invalid_bin2, set) %>%
	mutate(outcome=as.numeric(bxy != 0)) %>%
	summarise(
		n=n(),
		auc = roc(outcome, -log10(pval))$auc,
		psig=sum(pval < 1e-3, na.rm=TRUE) / n(),
		bias=sum(pdiff < 0.05, na.rm=TRUE) / n(),
		isq=mean(Isq)
	)

p2 <- sumu1b %>% 
filter(!grepl("Hybrid", method)) %>%
ggplot(aes(y=auc, x=prop_invalid_bin2, group=as.factor(method)), colour=as.factor(method)) +
geom_point(aes(colour=as.factor(method))) +
geom_line(aes(colour=as.factor(method))) +
scale_colour_brewer(type="qual") +
labs(x="Proportion of SNPs with pleiotropic effects", y="AUROC", colour="Method", shape="Outlier detection", linetype="Outlier detection") +
ylim(c(0.45, 1))
p2
```


## U2

The influence of invalid instruments on the performance of each method in terms of **bias**

```{r}
simresu2 <- subset(simres, set == "U2")
simresu2$directional_bias2 <- simresu2$directional_bias
simresu2$directional_bias2[simresu2$directional_bias] <- "Directional"
simresu2$directional_bias2[!simresu2$directional_bias] <- "Balanced"
sumu2 <- simresu2 %>% 
	group_by(method, prop_invalid_bin2, set, directional_bias2) %>%
	summarise(
		n=n(),
		psig=sum(pval < 1e-6, na.rm=TRUE) / n(),
		bias=sum(pdiff > 0.05, na.rm=TRUE) / n(),
		isq=mean(Isq)
	) 

p3 <- sumu2 %>% 
filter(!grepl("Hybrid", method), !is.na(prop_invalid_bin2)) %>%
ggplot(aes(y=bias, x=prop_invalid_bin2, group=as.factor(method)), colour=as.factor(method)) +
geom_point(aes(colour=as.factor(method))) +
geom_line(aes(colour=as.factor(method))) +
facet_grid(. ~ directional_bias2) +
scale_colour_brewer(type="qual") +
labs(x="Proportion of SNPs with pleiotropic effects", y="Proportion of estimates that are unbiased", colour="Method", shape="Outlier detection", linetype="Outlier detection")
p3
```

Influence on power / false discovery rates. Calculate the AUROC for each method across the simulations

```{r}
sumu2b <- simresu2 %>% 
	subset(!is.na(prop_invalid_bin2)) %>%
	group_by(method, prop_invalid_bin2, set, directional_bias2) %>%
	mutate(outcome=as.numeric(bxy != 0)) %>%
	summarise(
		n=n(),
		auc = roc(outcome, -log10(pval))$auc,
		bias=sum(pdiff < 0.05, na.rm=TRUE) / n(),
		isq=mean(Isq)
	)

p4 <- sumu2b %>% 
filter(!grepl("Hybrid", method), !is.na(prop_invalid_bin2)) %>%
ggplot(aes(y=auc, x=prop_invalid_bin2, group=as.factor(method)), colour=as.factor(method)) +
geom_point(aes(colour=as.factor(method))) +
geom_line(aes(colour=as.factor(method))) +
facet_grid(. ~ directional_bias2) +
scale_colour_brewer(type="qual") +
labs(x="Number of SNPs with pleiotropic effects", y="AUROC", colour="Method", shape="Outlier detection", linetype="Outlier detection") +
ylim(c(0.45, 1))
p4
```

## U3


```{r}
simresu3 <- subset(simres, set == "U3")
simresu3$mediation <- "Partial mediation"
simresu3$mediation[simresu3$bxy == 0] <- "Complete mediation"
simresu3$bxy <- simresu3$bxy + simresu3$bxu3 * simresu3$bu3y
simresu3$diff <- simresu3$b - simresu3$bxy
simresu3$pdiff <- pt(abs(simresu3$diff)/simresu3$se, simresu3$nsnp, lower.tail=FALSE)

sumu3 <- simresu3 %>% 
	group_by(method, ngxu3, ngx, set, mediation) %>%
	summarise(
		n=n(),
		psig=sum(pval < 0.05, na.rm=TRUE) / n(),
		bias=sum(pdiff > 0.05, na.rm=TRUE) / n(),
		isq=mean(Isq)
	)

inMap <- function(x, from, to) (x - min(x)) / max(x - min(x)) * (to - from) + from

sumu3$psig <- inMap(c(0,1,sumu3$psig), 0.5, 1)[-c(1,2)]

p5 <- sumu3 %>% 
filter(!grepl("Hybrid", method), !is.na(ngxu3)) %>%
ggplot(aes(y=bias, x=ngxu3/ngx, group=as.factor(method)), colour=as.factor(method)) +
geom_point(aes(colour=as.factor(method))) +
geom_line(aes(colour=as.factor(method))) +
facet_grid(. ~ mediation) +
scale_colour_brewer(type="qual") +
labs(x="Proportion of SNPs with pleiotropic effects", y="Proportion of estimates that are unbiased", colour="Method", shape="Outlier detection", linetype="Outlier detection")
p5
```

Influence on power.

```{r}
p6 <- sumu3 %>% 
filter(!grepl("Hybrid", method), !is.na(ngxu3)) %>%
ggplot(aes(y=psig, x=ngxu3/ngx, group=as.factor(method)), colour=as.factor(method)) +
geom_point(aes(colour=as.factor(method))) +
geom_line(aes(colour=as.factor(method))) +
facet_grid(. ~ mediation) +
scale_colour_brewer(type="qual") +
labs(x="Proportion of SNPs with pleiotropic effects", y="Proportion of estimates that are unbiased", colour="Method", shape="Outlier detection", linetype="Outlier detection") +
ylim(c(0.45, 1))
p6
```

## Summary

Overall, what proportion of times does each of the methods best perform?



```{r}

overall <- bind_rows(
	sumu1 %>% filter(!grepl("Hybrid", method)) %>%
	group_by(prop_invalid_bin2) %>% 
	arrange(desc(bias)) %>%
	mutate(rank=1:n(), what="bias") %>%
	select(method, set, rank, what),

	sumu1b %>% filter(!grepl("Hybrid", method)) %>%
	group_by(prop_invalid_bin2) %>% 
	arrange(desc(auc)) %>%
	mutate(rank=1:n(), what="auroc") %>%
	select(method, set, rank, what),
	
	sumu2 %>% filter(!grepl("Hybrid", method)) %>%
	group_by(prop_invalid_bin2, directional_bias2) %>% 
	arrange(desc(bias)) %>%
	mutate(rank=1:n(), what="bias") %>%
	select(method, set, rank, what),

	sumu2b %>% filter(!grepl("Hybrid", method)) %>%
	group_by(prop_invalid_bin2, directional_bias2) %>% 
	arrange(desc(auc)) %>%
	mutate(rank=1:n(), what="auroc") %>%
	select(method, set, rank, what),

	sumu3 %>% filter(!grepl("Hybrid", method)) %>%
	group_by(ngxu3, mediation) %>% 
	arrange(desc(bias)) %>%
	mutate(rank=1:n(), what="bias") %>%
	select(method, set, rank, what),

	sumu3 %>% filter(!grepl("Hybrid", method)) %>%
	group_by(ngxu3, mediation) %>% 
	arrange(desc(psig)) %>%
	mutate(rank=1:n(), what="auroc") %>%
	select(method, set, rank, what)
) %>% select(method, set, rank, what)
```

How many scenarios are there in total?

```{r}
overall %>% group_by(method) %>% nest %>% slice(1) %>% unnest %>% nrow
```


Overall best ranked

```{r}
overall %>% group_by(method) %>% summarise(rank=sum(rank)) %>% arrange(rank) %>% kable
```

Best sensitivity and specififity

```{r}
overall %>% subset(what == "auroc") %>% group_by(method) %>% summarise(rank=sum(rank)) %>% arrange(rank) %>% kable
```

Most unbiased

```{r}
overall %>% subset(what == "bias") %>% group_by(method) %>% summarise(rank=sum(rank)) %>% arrange(rank) %>% kable
```



## Final plot


```{r}
library(gridExtra)
th1 <- theme_bw() + theme(legend.position="none", axis.title=element_blank(), axis.ticks.y=element_blank(), axis.text.y=element_blank())
th2 <- theme_bw() + theme(legend.position="none", axis.title.x=element_blank())

pdf("../images/allplots.pdf", width=12)
grid.arrange(
	p1+th2 + ylim(c(0,1)) + facet_wrap(~ ""), 
	p3+th1 + ylim(c(0.45,1)), 
	p5+th1 + ylim(c(0,1)), 
	p2+th2 + ylim(c(0.45,1)) + facet_wrap(~ ""), 
	p4+th1 + ylim(c(0,1)), 
	p6+th1 + ylim(c(0.45,1)) + theme(legend.position=c(0.75,0.4), legend.background=element_blank(), legend.text=element_text(size=6)) + labs(colour=""),
	layout_matrix=matrix(c(1,1,2,2,3,3,4,4,5,5,6,6), ncol=2) %>% t)
dev.off()

```


Make a table of all results

```{r}
library(tidyr)

dbias <- tibble()
dauc <- tibble()

dbias <- bind_rows(dbias, 
	sumu1 %>% 
	select(method, prop_invalid_bin2, bias) %>%
	spread(key=method, value=bias) %>% 
	mutate(scenario = "Genetic confounder")
)

dauc <- bind_rows(dauc, 
	sumu1b %>% 
	select(method, prop_invalid_bin2, auc) %>%
	spread(key=method, value=auc) %>% 
	mutate(scenario = "Genetic confounder")
)

dbias <- bind_rows(dbias, 
	ungroup(sumu2) %>% 
	filter(!is.na(prop_invalid_bin2), directional_bias2 == "Balanced") %>%
	select(method, prop_invalid_bin2, bias) %>%
	spread(key=method, value=bias) %>% 
	mutate(scenario = "Balanced horizontal pleiotropy")
)

dbias <- bind_rows(dbias, 
	ungroup(sumu2) %>% 
	filter(!is.na(prop_invalid_bin2), directional_bias2 == "Directional") %>%
	select(method, prop_invalid_bin2, bias) %>%
	spread(key=method, value=bias) %>% 
	mutate(scenario = "Directional horizontal pleiotropy")
)

dauc <- bind_rows(dauc, 
	ungroup(sumu2b) %>% 
	filter(!is.na(prop_invalid_bin2), directional_bias2 == "Balanced") %>%
	select(method, prop_invalid_bin2, auc) %>%
	spread(key=method, value=auc) %>% 
	mutate(scenario = "Balanced horizontal pleiotropy")
)

dauc <- bind_rows(dauc, 
	ungroup(sumu2b) %>% 
	filter(!is.na(prop_invalid_bin2), directional_bias2 == "Directional") %>%
	select(method, prop_invalid_bin2, auc) %>%	
	spread(key=method, value=auc) %>% 
	mutate(scenario = "Directional horizontal pleiotropy")
)

dbias <- bind_rows(dbias, 
	ungroup(sumu3) %>% 
	mutate(prop_invalid_bin2 = ngxu3/ngx) %>%
	filter(!is.na(prop_invalid_bin2), mediation == "Complete mediation") %>%
	select(method, prop_invalid_bin2, bias) %>%
	spread(key=method, value=bias) %>% 
	mutate(scenario = "Complete mediation")
)

dbias <- bind_rows(dbias, 
	ungroup(sumu3) %>% 
	mutate(prop_invalid_bin2 = ngxu3/ngx) %>%
	filter(!is.na(prop_invalid_bin2), mediation == "Partial mediation") %>%
	select(method, prop_invalid_bin2, bias) %>%
	spread(key=method, value=bias) %>% 
	mutate(scenario = "Partial mediation")
)

dauc <- bind_rows(dauc, 	
	ungroup(sumu3) %>% 
	mutate(prop_invalid_bin2 = ngxu3/ngx) %>%
	filter(!is.na(prop_invalid_bin2), mediation == "Complete mediation") %>%
	select(method, prop_invalid_bin2, psig) %>%
	spread(key=method, value=psig) %>% 
	mutate(scenario = "Complete mediation")
)

dauc <- bind_rows(dauc, 
	ungroup(sumu3) %>% 
	mutate(prop_invalid_bin2 = ngxu3/ngx) %>%
	filter(!is.na(prop_invalid_bin2), mediation == "Partial mediation") %>%
	select(method, prop_invalid_bin2, psig) %>%	
	spread(key=method, value=psig) %>% 
	mutate(scenario = "Partial mediation")
)

write.csv(dauc, file="../results/sim_auc.csv")
write.csv(dbias, file="../results/sim_bias.csv")
